
---
title: "Analysis Report for Reproducibility Project"
output: html_document
date: "2025-02-23"
---

```{r setup, include=FALSE}
# Set up chunk options globally for the entire document
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose of this Document

This markdown file allows you to easily run the code that generates plots for the 2025 honors reproducibility project. The necessary packages for the analysis are listed below.

```{r load_packages, echo=FALSE}
# Load the required libraries for the analysis
library("tidyverse")
library("patchwork")
```

## Loading and Cleaning the Data

The following code transforms the original data into long format and creates a new variable called `Species`, which includes columns 4 to 49 from the original database.

```{r load_data}
# Load the original data
odat <- read_csv("YOYGroundfishBiodiv_Data.csv")

# Define the column names for species data
species_columns <- colnames(odat[4:49])

# Transform the data into long format
odat_long <- odat %>%
  select(-YEAR_STATION) %>%
  pivot_longer(
    cols = all_of(species_columns),
    names_to = "Species",
    values_to = "Taxon_CPUE(log + 1)"
  )
```

## Preparing Data for Plots

The code below prepares the data by summarizing CPUE and Species Diversity Index (sdiv) for 2023 and 1990. The data is reordered by the median CPUE and sdiv values.

### Un-Summarized Data for 2023 (CPUE and sdiv by Area)

```{r data_2023}
# Filter and organize 2023 data
sd_23_unsum <- odat_long %>% 
  filter(YEAR == "2023") %>%
  select(AREA, cpue, sdiv) %>%
  group_by(AREA) %>%
  arrange(cpue)

# Reorder areas by median CPUE
sd_23_cpue_ordered <- sd_23_unsum
sd_23_cpue_ordered$AREA <- reorder(sd_23_cpue_ordered$AREA, sd_23_cpue_ordered$cpue, FUN = median, decreasing = TRUE)

# Reorder areas by median sdiv
sd_23_sdiv_ordered <- sd_23_unsum
sd_23_sdiv_ordered$AREA <- reorder(sd_23_sdiv_ordered$AREA, sd_23_sdiv_ordered$sdiv, FUN = median, decreasing = TRUE)
```

### Un-Summarized Data for 1990 (CPUE and sdiv by Area)

```{r data_1990}
# Filter and organize 1990 data
sd_90_unsum <- odat_long %>% 
  filter(YEAR == "1990") %>%
  select(AREA, cpue, sdiv) %>%
  group_by(AREA) %>%
  arrange(cpue)

# Reorder areas by median CPUE
sd_90_cpue_ordered <- sd_90_unsum
sd_90_cpue_ordered$AREA <- reorder(sd_90_cpue_ordered$AREA, sd_90_cpue_ordered$cpue, FUN = median, decreasing = TRUE)

# Reorder areas by median sdiv
sd_90_sdiv_ordered <- sd_90_unsum
sd_90_sdiv_ordered$AREA <- reorder(sd_90_sdiv_ordered$AREA, sd_90_sdiv_ordered$sdiv, FUN = median, decreasing = TRUE)
```

## Plotting the Data

### Boxplots for CPUE and Species Diversity Index for 2023 and 1990

```{r plot_2023}
# Plot for CPUE (2023)
bp1_avcpue_23 <- ggplot(sd_23_cpue_ordered, aes(x = AREA, cpue)) +
  geom_boxplot(fill = "slateblue", alpha = 0.2) +
  labs(x = "AREA", y = "CPUE") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(panel.grid.major = element_blank())

# Plot for Species Diversity Index (2023)
bp2_avsdiv_23 <- ggplot(sd_23_sdiv_ordered, aes(x = AREA, sdiv)) +
  geom_boxplot(fill = "slateblue", alpha = 0.2) +
  labs(x = "AREA", y = "Species Diversity Index") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(panel.grid.major = element_blank())
```

```{r plot_1990}
# Plot for CPUE (1990)
bp3_avcpue_90 <- ggplot(sd_90_cpue_ordered, aes(x = AREA, cpue)) +
  geom_boxplot(fill = "slateblue", alpha = 0.2) +
  labs(x = "AREA", y = "CPUE") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(panel.grid.major = element_blank())

# Plot for Species Diversity Index (1990)
bp4_avsdiv_90 <- ggplot(sd_90_sdiv_ordered, aes(x = AREA, sdiv)) +
  geom_boxplot(fill = "slateblue", alpha = 0.2) +
  labs(x = "AREA", y = "Species Diversity Index") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(panel.grid.major = element_blank())
```

### Displaying the Plots for 2023 and 1990

```{r display_plots}
# Combine 2023 plots (CPUE and sdiv)
bp1_avcpue_23 + bp2_avsdiv_23  # 2023 CPUE and Species Diversity Index

# Combine 1990 plots (CPUE and sdiv)
bp3_avcpue_90 + bp4_avsdiv_90  # 1990 CPUE and Species Diversity Index
```

## Species Diversity Index per Year

```{r data_species_diversity_per_year}
# Prepare the data for species diversity index per year
sd_year <- odat_long %>% 
  select(YEAR, sdiv) %>%
  group_by(YEAR)

# Create the plot for species diversity index per year
avg_sdiv_year <- ggplot(sd_year, aes(x = YEAR, sdiv, group = YEAR)) +
  geom_boxplot(fill = "slateblue", alpha = 0.2) +
  labs(x = "YEAR", y = "Species Diversity Index") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(panel.grid.major = element_blank())
```

### Plot for Species Diversity Index per Year

```{r display_avg_sdiv_year, echo=FALSE}
# Display the species diversity index per year plot
avg_sdiv_year
```

